% !TEX program=pdflatex

\documentclass[x11names]{article}

\usepackage[style=authoryear]{biblatex}
\addbibresource{zugzwang.bib}

\input{macros}
\usepackage{geometry}
\geometry{
    paper=a4paper,
    left=1cm,
    right=5cm,
    top=1cm,
    bottom=2cm
}

\title{Weighted Answer Set Programs}
\author{Francisco, Bruno, Salvador, Dietmar}

\begin{document}

\maketitle

\begin{abstract}
    Drawing inspiration from HMMs; State of the art of PLP and Probabilistic ASP; Leveraging current Prob ASP systems; 
\end{abstract}

Using a \acl{LP} to model and reason over a real world situation is often complicated because of uncertainty underlying the problem being worked on.
Classic \aclp{LP} represent knowledge in precise terms, which turn out to be problematic when the application data is affected by stochastic factors, which is frequently the case.

\section{Uncertainty and ASPs}


\begin{figure}[t]
	\begin{center}
		\begin{tikzpicture}[node distance=4em]
            \node[boxed, fill=cyan!5] (A) {system};
			\node[boxed, right = of A, fill=cyan!5] (S) {state};
			\node[boxed, right = of S, fill=cyan!5] (z) {observation};
			\node[boxed, below = of A, fill=yellow!5] (P) {program};
			\node[boxed, right = of P, fill=yellow!5] (m) {stable model};
			\node[boxed, right = of m] (L) {literals};
			% ----
            \path[-latex] 
                (A) edge node[above] {\emph{is in a\ldots}} (S)
                (P) edge node[left] {\emph{models}} (A)
                (S) edge node[above] {\emph{reveals}} (z)
                (S) edge[dotted,-] node[right] {\emph{coincide}} (m)
                (L) edge[bend right, dashed] node[right] {\emph{describe}} (z)
                (L) edge[dashed] node[above]{\emph{describe}} (m)
                (P) edge node[below] {\emph{entails}} (m);
		\end{tikzpicture}
	\end{center}

	\caption{%
		}
\end{figure}

\Acp{HMM} setup a well known, widely used, framework to describe and study partially observable stochastic processes\sidenote{HMMs stress the \textbf{process} and we are not looking into that.}. Briefly, a \ac{HMM} process is described by a sequence of states, each defined as a set of random variables that are partially observed. This means that (1) observation of states include only a fixed subset of those variables, and also that (2) the observed values are subject to random perturbations.

We try to capture both the partial observability and the stochastic nature of a system in a logic setting. To this end we think of a system, $S$, as a formal model, $P$, together with a set of observations, $D$:

\begin{equation}\label{eq:def.system}
    S = \left( P, D \right)
\end{equation}

The formal model is an \acl{ASP} extended with \emph{weights}, that we describe later in this paper, and such that the \aclp{SM} of the program are the system states. On the other hand, any given system state can be (or has been) observed. The outcome of each single observation is a set of literals of that program.

\begin{example}[Coins, Dices, and Cards]
    Consider a scenario where a coin is tossed and, if the result is \emph{heads}, \(a\), then a dice is thrown, \(b\), or a card is drawn, \(c\). If the coin lands in \emph{tails}, \(\neg a\), no information is given. A formal model of this system is the program  
    \begin{equation}\label{eq:derived.fruitful}
        P_{\sbf} = \left\lbrace\begin{aligned}
            a \vee \neg{a}  & ,          %
            \\
            b \vee c        & \clause a
        \end{aligned}
        \right.
    \end{equation}
    that has atoms \(\set{a, b, c}\), literals \(\set{a, \neg a, b, \neg b, c, \neg c}\) and \aclp{SM}
    \begin{equation}
        \set{a, b}, \set{a, c}, \set{\neg a}.
    \end{equation}
    
    Possible observations include:
    \begin{itemize}
        \item \(\set{a,c}\): a \ac{SM}, therefore a system state, observed \emph{heads} and \emph{card}.
        \item \(\set{a}\): not a \ac{SM} but contained in the two \acp{SM} $\set{a, b}$ and $\set{a, c}$, observed \emph{heads} but we have no information about the \emph{cards} or the \emph{dice}.
        \item \(\set{\neg a, \neg b}\): not a \ac{SM} but contains the \ac{SM} $\set{\neg a}$, the coin was observed as \emph{not heads} (\emph{tails}?); also we are certain that no card was drawn but have no information about the dice.
        \item \(\set{b,c}\): not related with any \ac{SM}.
        \item \(\set{a,\neg a}\): an inconsistent observation.
    \end{itemize}    
\end{example}

We use sets of literals, instead of atoms, to represent observations because they make possible the distintion between a \emph{explicit negative observation} \(\neg\alpha\) (\eg ``I see tails.'') and \emph{not observed} \(\naf \alpha\) (\eg ``The coin in hidden.''). This corresponds to assuming a ``boolean sensor'' for each $\alpha$ and $\neg \alpha$.

So, the program \ref{eq:derived.fruitful} defines the ``sensors''
\begin{equation*}
    a, \neg a, b, \neg b, c, \neg c
\end{equation*}

\nocite{*}

\printbibliography
\end{document}